Building Markov State Models for Periodically Driven Non-Equilibrium Systems

### 2014-11-05 writedown/period.noneq.msm/paper.tex
merge CS's revise, add some from mine

\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}
\newcommand{\bluec}[1]{{\color{red} #1}}
\newcommand{\greenc}[1]{{\color{green} #1}}
% \newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\vect}[1]{#1}
\newcommand{\dd}[1]{\textsf{#1}}
% \newcommand{\fwd}[0]{\textrm{fw}}
% \newcommand{\bwd}[0]{\textrm{bw}}
\newcommand{\fwd}[0]{+}
\newcommand{\bwd}[0]{-}
\newcommand{\period}[0]{T_{\textrm{P}}}
\newcommand{\ml}[0]{\mathcal {L}}
\newcommand{\mo}[0]{\mathcal {O}}
\newcommand{\mbp}[0]{\mathbb {P}}
\newcommand{\mc}[0]{\mathcal {C}}
\newcommand{\dt}[0]{\Delta t}
\newcommand{\id}{\mathrm{Id}}
% \newcommand{\myphi}{\boldsymbol\Phi}
% \newcommand{\mymu}{\boldsymbol\mu}
\newcommand{\myphi}{\Phi}
\newcommand{\mymu}{\mu}
\newcommand{\prob}{\textrm{P}}
\newcommand{\dih}{\textrm{dih}}

@@

\title{Building Markov State Models for Periodically Driven Non-Equilibrium Systems}
\author{Han Wang}
\email{han.wang@fu-berlin.de}
\affiliation{Zuse Institut Berlin, Germany}
\author{Christof Sch\"utte}
\email{Christof.Schuette@fu-berlin.de}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\affiliation{Zuse Institut Berlin, Germany}

\begin{abstract}
\end{abstract}

@@

The non-equilibrium biomolecular systems, especially
those systems under periodic driven, are attracting increasing
interest.  For example, the potential effects of an electromagnetic
radiation on the human body tissue (e.g.~DNA, protein, and membrane)  has been extensively
investigated by a vast number of literatures, and a very limited
subset of them is Ref.~\cite{bohr2000microwave, bohr2000microwave-1, dePomerai2000cell,
  dePomerai2003microwave, mancinelli2004non, inskip2001cellular, bekard2013electric, budi2005electric,
  budi2007effect, budi2008comparative, astrakas2012structural,
  damm2012can, english2009nonequilibrium, solomentsev2012effects,
  wang2014exploring}.  Among those theoretical investigations, the
molecular dynamics (MD) has been shown  useful in understanding
the response of the biomolecular conformation to the external electric field,
due to its ability of resolving the molecular
details, which are sometimes not achievable in experiments. Only
recently, a dynamical non-equilibrium MD simulation (D-NEMD) method has
been proposed.  With the local thermostating technique, the D-NEMD
method is able to sample the non-equilibrium probability density at the
constant temperature, therefore opens the possibility of studying the
non-equilibrium processes in a statistically reliable and
thermodynamically consistent way.

@@

Despite their high resolution in presenting a molecular system, MD the
trajectories including the time-dependent position and velocity of every atom
sometimes impose non-trivial challenge for data analyzing, in the sense that 
extracting the essential structure and dynamical properties is
becoming increasingly difficult as the  system size and trajectory length grow.
% by merely visulizing (not only molecular cartoons but also order parameters)
% and observing the molecules.
The Markov state model
(MSM)~\cite{A19-31}, which is a promising method for data analyzing of
the high-dimensional MD trajectories, has been well developed over the
past decade in theory and applications \cite{A19-31, A19-1}, and in
the software implementations~\cite{A19-49, MSMBuilder}.  The principal
idea of MSM is to approximate the original high-dimensional system by
a reduced Markovian dynamics over a finite number of states, from
which the essential structures and dynamics can be formulated by the spectrum analysis.
By assigning the spacial and temporal discretizations, 
the transition probabilities are sampled from MD trajectories, 
% the MSM does not need more information from the MD simulation.
so the quality of the MSM  relies on the quality of the discretizations.
Moreover, the characteristic of the original dynamics plays a role:
It has been shown that in principle good approximate accuracy would be achieved
by using only a few ``coresets'',
if the system is dominated by a few metastable states, in the sense that
the MD trajectories stay in the vicinity of a metastable state substantially longer than
transform to another such state~\cite{schuette2011markov}.
Besides the applications in analyzing the MD trajectories, it worth noting that MSM is
also used for designing initial configurations for massively parallel MD
simulations~\cite{kohlhoff2014cloud}.

@@

% \textbf{Importance: first application of MSM in a non-equilibrium system.}

Despite the significance of the non-equilibrium phenomena, the
analysis of the non-equilibrium MD simulations mainly follows the
visualizing (not only molecular cartoons but also order parameters)
and observing approach, and the quantitative description of
the essential conformational dynamics is still lacking. On the other
hand, the MSM, to the best of our knowledge, has only been applied to
equilibrium systems,  and most of them are assumed reversible. This work is the first attempt
of using the MSM in analyzing a class of truly non-equilibrium systems:
periodic external driven. Being more specifically,
we investigate the protein conformational dynamics
under the oscillating electric field (EF).
In Sec.~\ref{sec:disc}, we discuss the temporal and spacial
discretizations of the non-equilibrium dynamics. The spacial discretization
is on the dihedral angle space, which can be interpreted as a
MSM in the traditional sense of full spacial partition.
In the temporal direction, the non-equilibrium process
is  discretized by the Floquet theorem into a time-homogeneous
but not necessarily reversible dynamics. Since the full spacial partition
is not preferable in high-dimensional system, in Sec.~\ref{sec:build-msm},
the MSM based on milestoning is introduced for the discretized irreversible process.
The validity of the discretizations and the MSM is checked in Sec.~\ref{sec:alanine}
by comparing to the brute force
non-equilibrium simulation of alanine dipeptide under oscillating EF.
This work is summarized in Sec.~\ref{sec:conclusion} by a chart presenting
the workflow of building MSM for non-equilibrium systems,
and open questions are presented in the end.

@@

\section{Non-equilibrium molecular dynamics and its discretization}
\label{sec:disc}
For the sake of simplicity we consider diffusive molecular dynamics in an energy landscape $V$ driven by the time-dependent external driving force $E(t)D(x_t)$ with the
$T$-periodic external field $E(t)$:
\begin{align}
  \label{eq:disc-1}
  d\vect x_t = \Big(-\nabla V(\vect x_t) + E(t) D(\vect x_t)\Big)dt + \sqrt{2\beta^{-1}} d\vect w_t, 
\end{align}
where $x_t\in\Omega$ denotes the state of the molecular system at time $t$ in state space $\Omega$, $\vect w_t$ denotes standard $n$-dimensional Brownian motion, and $\beta$ the inverse temperature,
i.e.~$\beta = 1/(k_B\mathcal T)$.
The propagation of probability
densities $\rho=\rho(\vect x,t)$ based on this kind of dynamics in the sense
of $\rho(\vect x,t)dx=\prob[\vect x_t\in [\vect x,\vect x+d\vect x)]$ is governed by
Fokker-Planck equation:
\begin{align}
  \label{eq:disc-fp}
  \frac{\partial \rho}{\partial t} = \ml^\dagger(t) \rho,
\end{align}
where $\ml^\dagger(t)$ is the adjoint of the generator
\begin{align}
  \label{eq:disc-3}
  \ml(t)=\beta^{-1}\Delta_{\vect x}+\Big(-\nabla_x V(\vect x) + E(t)D(\vect x)\Big)\cdot\nabla_{\vect x},
\end{align}
where $\Delta_{\vect x}$ denotes the Laplacian operator and $\nabla_{\vect x}$
the nabla-operator wrt to $\vect x$. 
The periodicity of the external driving force induces the periodicity of the generator,
i.e.~$\ml(t) = \ml(t+T)$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial discretization: Master equation}
We will now introduce an appropriate spatial discretization of this kind of non-equilibrium MD -- this is done for reasons of simplicity only; we could completely avoid it for the price of more technical arguments.
For achieving this discretization, we 
introduce a partition of state space $\Omega$ into a finite number of disjoint
sets $\{ \Omega_1, \cdots, \Omega_N\}$ satisfying $\Omega = \cup_i \Omega_i$,
$\Omega_j\cap \Omega_j = \emptyset,\ \forall i\neq j$.
Utilizing the procedure described in Ref.~\cite{latorre2011structure} the original Fokker-Planck equation~(\ref{eq:disc-fp})
is discretized, resulting in a time-inhomogeneous Markov jump process in state
space $S = \{1, \cdots, N\}$ with time-dependent rate
matrix $\vect L(t) \in \mathbb R^{N\times N}$ satisfying
\begin{align}\label{eq:disc-4}
\sum\limits_{j=1}^N L_{ij}(t) & =  0\\ \label{eq:disc-5}
L_{ij}(t) & \ge  0, \quad i\not= j\\
L_{ij}(t) & =  L_{ij}(t+T)
\end{align}
for all real time $t\geq 0$.
Moreover, the rate matrix $L$ has the form $\vect L(t)=\vect L_0+E(t)\vect L_1$
where $E(t)$ is periodic with period $T>0$.
In analogy to \eqref{eq:disc-fp}, the Markov jump process generated by
$\vect L(t)$ transports probability distributions according to the associated Master equation
\begin{align}
  \label{eq:disc-master}
  \frac{d\vect p(t)}{dt} = \vect L^{\top}(t)\cdot \vect p(t)
\end{align}
where $\vect L^{\top}(t)$ denotes the matrix transpose of $\vect L(t)$, $\vect p(t)$ is an $N$-vector denoting the probability distribution on $S$ at time $t$, $p(i,t)$, for example, the probability to be in state $i$ (which corresponds to set $\Omega_i$) at time $t$.
As usual the properties (\ref{eq:disc-4}) and (\ref{eq:disc-5}) of
$\vect L(t)$ guarantee that the total probability mass is conserved,
i.e., if $p(i,0)\ge 0$ componentwise, then $p(i,t)\ge 0$ and $\sum_i
p(i,t) = \sum_ip(i,0)$.
% The solution of the master equation need no be
% periodic.
The temporal evolution of the probability distribution $\vect p(t)$ can be formally written
\begin{align}  \label{eq:disc-8}
\vect p(t)=\myphi(t)\vect p(0)
\end{align}
by using the
associated propagator matrix $\myphi(t)\in\mathbb R^{N\times N}$ that solves
\begin{align}
  \label{eq:disc-master-phi}
  \frac{d}{dt}\myphi(t) = \vect L^{\top}(t)\myphi(t), \quad \myphi(0) = \id .
\end{align}
Since the last equation can be considered column-wise, the propagator matrix inherits column-wise conservation properties:
$\Phi_{ij}(t) \ge  0$
and $\sum\limits_{i=1}^N \Phi_{ij}(t)  =  1$,
that is, $\myphi^{\top}(t)$ is a
stochastic matrix satisfying $\myphi^{\top}(t)\vect e=\vect e$
with $\vect e=(1,\ldots,1)^{\top}\in \mathbb R^N$.
Regarding these considerations, we find
\begin{align}
\label{eq:disc-10}  
\myphi_{ij}(t)=\prob\left(\vect X_t=i\mid \vect X_0=j \right),
\end{align}
where $\vect X_t$ denotes the Markov process generated by $\vect L(t)$.

@@

Using the Floquet theorem, the time-inhomogeneous Markov jump process $X_t$
is therefore discretized into a \emph{time-homogeneous} (not necessarily
reversible) Markov jump process $\tilde X_{m} = X_{mT}, \ m\in\mathbb
N$, which is generated by transition matrix
\begin{equation}\label{P}
\vect P=\myphi^{\top}(T).
\end{equation}
We prefer to consider the discrete-time process $\tilde X_{m}$ instead of the time-continuous process $X_t$
because the powerful theories and
computational tools for time-homogeneous Markov processes can be directly applied. It worth noting that many of these tools require the transition matrix $P$ to satisfy the detailed balance condition. The computations in the Appendix~\ref{sec:app-revs} show that $P$ will in general \emph{not} satisfy this condition; in fact  the deviation form reversibility can be estimated from the work of the periodic driving does to the system.
There is no doubt that information within one period is lost by using this temporal
discretization, however, information regarding the long-term behavior of the system on time-scales
much longer than the period will be perfectly described because of $\tilde X_{m} = X_{mT}\approx x_{mT}$ whenever our spatial discretization is fine enough.
At the same time,
the computational cost of generating $\tilde X_{m}$ is much less demanding
than the brute force simulations of NEMD, which implies lower
statistically uncertainty in calculating the observables of interest.

Since $P$ is a stochastic matrix, its eigenvalues are contained in the unit circle in
the complex plane, i.e., each eigenvalue $\lambda$ (potentially complex-valued)
satisfies $|\lambda|\le 1$. Furthermore $\lambda=1$ is an
eigenvalue with right eigenvector $\vect e=(1,\ldots,1)^{\top}$ and a left eigenvector $\mymu$
satisfying
$\mymu^{\top}P=\mymu^{\top}$.
From now on, we assume $P$ to be irreducible and aperiodic such that the Perron Frobenius theorem the eigenvalue $\lambda=1$ is  non-negative componentwise, and unique (up to normalization $\sum_i\mu(i)=1$). In this case $\mymu$ is the stationary measure in the sense that
$
 \mymu^{\top} P^m = \mymu^{\top},\ m\in\mathbb N
$,
% \begin{align}
% \label{eq:floq-14}  
% \end{align}
and (more precisely) the asymptotic evolution of an initial probability distribution $\vect p(t=0)$ by the process satisfies
$
\vect p^{\top}(0)\,P^m \to \mymu^{\top},\ m\to\infty,
$
% \begin{align}
% \label{eq:floq-15}  
% \end{align}
so that $\mymu$ can be seen as the quasi-stationary distribution of our non-stationary process.

% due to the discretization in both the spacial and temperal directions.
% In the numercal example in Sec.~\ref{sec:alanine}, we
% compute the

@@

% Given coresets $\{C_1, \cdots, C_k\}$, the procedure \eqref{eq:msm-ht}--\eqref{eq:msm-tmatrix-10}
% If assuming that we have the coresets $\{C_1, \cdots, C_k\}$, no matter how they are identified,

@@

We know that --in theory-- whenever the spatial discretization is fine enough, the Markov jump
process $X_t$ associated with the Master equation \eqref{eq:disc-master} is a good approximation to the original
MD process $x_t$ governed by \eqref{eq:disc-1}.  In practice, however, it is difficult to predict
how many discrete sets we need to be fine enough. Moreover, since the
total dimension of $x_t$ is $3N_{\textrm{atom}}$ ($N_{\textrm{atom}}$ being the number of atoms), it is prohibitive to do a really fine discretization
over all DOFs for most systems of practical interest.

One possible way to define appropriate discretization sets is firstly to find a few collective
variables, and then to discretize these collective variables as finely as needed either by uniform or adaptive
discretization~\cite{chodera2007automatic, prinz2011markov}.
% This process also suggests that the choice of the lag-time should be 
% shorter than the dominant implied timescale so that they can be resolved,
% and be longer than the fast time-scales to relax the unresolved dynamics
% within the discretized sets.
However, it is difficult to give a general answer in prior regarding
how to choose the collect variables and how fine their discretization should be.
For large or high dimensional systems, these questions usually become non-trivial.

@@

An important check is the consistency between the
stationary probability density of $\Phi(T)$ (i.e.~the leading eigenvector $\mymu$) and that 
estimated form the original NEMD simulation,
% along which only integral periods $mT$ are considered:
\begin{align}
  \label{eq:num-tmp1}
  \rho_{\textrm{st}}(\phi,\psi) = \lim_{m\rightarrow\infty} \rho (\phi,\psi,mT),
\end{align}
On each NEMD branching trajectory the initial 320~ps are discarded and
the rest of the trajectory in time interval $[320,4000]$~ps is used to estimate
the quasi-stationary probability distribution $\rho_{\textrm{st}}$. 
$P$ is computed as described above, and then $\mymu$ is computed as its leading eigenvector. In order to make it comparable to the free energy in the equilibrium case, we take
the logarithm of the distributions, i.e.~$F_{\textrm{st}}(\phi,\psi)=
-k_B\mathcal T\log \rho_{\textrm{st}}(\phi,\psi)$
for NEMD and $F_{\textrm{st}}(\phi,\psi)=
-k_B\mathcal T\log \mu(\phi,\psi)$ for $P$, where $k_B$ is the
Boltzmann constant and $\mathcal T$ is the temperature of the system.
% Under the aformentioned discretization, the stationary distribution is
% defined for the each bin of the dihedral angle space, i.e.~$F_{pq} =
% \int_{ph}^{(p+1)h}d\phi\int_{qh}^{(q+1)h}d\psi
% F_{\textrm{st}}(\phi,\psi), \ 0\leq p,q<N_\dih-1$, where $h = 360/N_\dih$ being
% the size of the bin, $p$ and $q$ here being the bin indexes.
The
results are  compared in Fig.~\ref{fig:num-1}. A good consistency between
the  NEMD simulation and 
$P$ is observed.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Core set identification}

>>>>

\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}
\newcommand{\bluec}[1]{{\color{blue} #1}}
\newcommand{\greenc}[1]{{\color{green} #1}}
% \newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\vect}[1]{#1}
\newcommand{\dd}[1]{\textsf{#1}}
% \newcommand{\fwd}[0]{\textrm{fw}}
% \newcommand{\bwd}[0]{\textrm{bw}}
\newcommand{\fwd}[0]{+}
\newcommand{\bwd}[0]{-}
\newcommand{\period}[0]{T_{\textrm{P}}}
\newcommand{\ml}[0]{\mathcal {L}}
\newcommand{\mo}[0]{\mathcal {O}}
\newcommand{\mbp}[0]{\mathbb {P}}
\newcommand{\mc}[0]{\mathcal {C}}
\newcommand{\dt}[0]{\Delta t}
\newcommand{\id}{\mathrm{Id}}
% \newcommand{\myphi}{\boldsymbol\Phi}
% \newcommand{\mymu}{\boldsymbol\mu}
\newcommand{\myphi}{\Phi}
\newcommand{\mymu}{\mu}
\newcommand{\prob}{\textrm{P}}
\newcommand{\dih}{\textrm{dih}}

@@

\title{Building Markov State Models for Periodically Driven Non-Equilibrium Systems}
\author{Han Wang}
\email{han.wang@fu-berlin.de}
\affiliation{CAEP Software Center for High Performance Numerical Simulation, Beijing, China}
\affiliation{Zuse Institute Berlin, Germany}
\author{Christof Sch\"utte}
\email{Christof.Schuette@fu-berlin.de}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\affiliation{Zuse Institute Berlin, Germany}

\begin{abstract}
Recent years have seen an increased interest in non-equilibrium molecular dynamics (NEMD) simulations, especially for molecular systems with periodic forcing by external fields, e.g., in the context of studying  effects of electromagnetic radiation on the human body tissue. Lately, an NEMD methods with local thermostating has been proposed that allows for studying 
non-equilibrium processes in a statistically reliable and thermodynamically consistent way. In this article, we demonstrate how to construct Markov State Models (MSMs) for such NEMD simulations. MSM building has been well-established for systems in equilibrium where MSMs with just a few (macro-)states allow for accurate reproduction of the essential kinetics of the molecular system under consideration. Non-equilibrium MSMs have been lacking so far. The article presents how to construct such MSMs and illustrates their validity and usefulness for the case of conformation dynamics of alanine dipeptide in an external electric field.  
\end{abstract}

@@

Biomolecular systems under non-equilibrium conditions caused by external fields, especially
systems under periodic forcing, have attracted increasing
interest recently.  For example, the potential effects of electromagnetic
radiation on the human body tissue (e.g.~DNA, protein, and membrane)  has been extensively
investigated in a vast number of articles, with the following list just representing an incomplete selection~\cite{bohr2000microwave, bohr2000microwave-1, dePomerai2000cell,
  dePomerai2003microwave, mancinelli2004non, inskip2001cellular, bekard2013electric, budi2005electric,
  budi2007effect, budi2008comparative, astrakas2012structural,
  damm2012can, english2009nonequilibrium, solomentsev2012effects}.  Molecular dynamics (MD) simulations have proved particularly useful for understanding
the response of biomolecular conformations to external fields because of their ability to  resolve molecular
details that \recheck{sometimes} cannot be resolved in experiments. Only
recently, a \recheck{dynamical} non-equilibrium MD simulation (D-NEMD) method with local thermostating has been proposed~\cite{wang2014exploring} that allows for studying 
non-equilibrium processes in a statistically reliable and thermodynamically consistent way. 
Despite the significance of the non-equilibrium phenomena, the
analysis of the non-equilibrium MD simulations mainly follows standard approaches, and reliable tools for quantitative description of
the essential conformational dynamics of the molecular system under external forcing are still lacking.

@@

Despite their many advantages, MD simulations have severe limitations. For example, one has to assume that the underlying force fields are appropriately describing the internal and external molecular interactions, and the maximal possible simulation length often is shorter than the timescale of interest. 
This article is mainly concerned with circumventing the latter obstacle by introducing non-equilibrium Markov State Models. Markov State Models
(MSM) have been well developed over the
past decade in theory~\cite{A19-31,prinz2011markov} and applications (see the recent book~\cite{A19-1} for an overview), and software implementations~\cite{A19-49, MSMBuilder}, but for systems under equilibrium conditions only!  
The principal idea of equilibrium MSMs is to approximate the original high-dimensional MD system by
a reduced Markovian dynamics over a finite number of (macro-)states. These (macro-)states
have to be identified with the dominant metastable sets, in the sense that
typical MD trajectories stay in the vicinity of a metastable set substantially longer than
the systems needs for a transition to another such state~\cite{A19-31,schuette2011markov}. In this case, the metastable sets are the main conformations of the molecular system under consideration which, often enough, are given by the main wells in the energy landscape. 
It has been shown that for molecular systems exhibiting such metastable sets the Markovian dynamics given by an MSM allows very close approximation of the longest relaxation processes of the underlying molecular system, at least under equilibrium conditions~\cite{sarich2010approximation,Eigenvalues}. 
Moreover, it has been demonstrated that in such cases MSM building requires short MD trajectories only, much shorter than the timescales of interest~\cite{PNAS09,kohlhoff2014cloud}.  
Thus, MSM building often allows to study the dynamical behavior on long timescales without requiring MD trajectories of comparable length.
Moreover, MSMs are utilized for understanding very long MD simulations: Extracting the essential structures and dynamical properties from long MD runs is
becoming increasingly difficult as the system size and trajectory length grow; MSMs have been used to construct kinetic fingerprints from MD simulations~\cite{A19-39} that help understanding the essential dynamics and in comparison with experiments~\cite{PrinzKellerNoe_PCCP11_Perspective}.

@@

% \textbf{Importance: first application of MSM in a non-equilibrium system.}
To the knowledge of the authors this work is the first attempt of using MSMs for analyzing non-equilibrium systems under periodic external forcing.
More precisely, we will demonstrate how to use MSMs for investigate the conformational dynamics of a peptide (alanine dipeptide) under an oscillating electric field (EF).
To this end, we will show how to generalize Markov state modeling to periodic non-equilibrium conditions where one cannot assume reversibility of the dynamics as it is mostly done in the literature on MSM building.

The outline of the article is as follows: 
In Sec.~\ref{sec:disc}, we discuss the temporal and spatial
discretizations needed to construct an MSM. We consider spatial discretization
of the dihedral angle space in the traditional sense of full partition MSMs~\cite{pande2010everything,A19-29}.
In the temporal direction, the non-equilibrium process
is discretized utilizing Floquet's theorem. This results in a time-homogeneous, 
but not necessarily reversible, that is, irreversible Markov process. Since  full spatial partition does not make any sense
for high-dimensional systems, we next show how to construct a \emph{few-state} MSM based on milestoning~\cite{schuette2011markov,A19-29} of the discretized irreversible Markov process in Sec.~\ref{sec:build-msm}.
The validity of the discretizations and the resulting MSM is checked in Sec.~\ref{sec:alanine}
by comparing the kinetic fingerprints given by the MSM to brute force
non-equilibrium \recheck{MD} simulations of alanine dipeptide under oscillating EF.
The findings are summarized in Sec.~\ref{sec:conclusion} including a chart presenting
the workflow of MSM building for non-equilibrium systems, and a list of open questions.

@@

\section{Non-equilibrium molecular dynamics and its discretization}
\label{sec:disc}
\recheck{We consider diffusive molecular dynamics} in an energy landscape $V$ driven by the time-dependent external driving force $E(t)D(x_t)$ with the
$T$-periodic external field $E(t)$:
\begin{align}
  \label{eq:disc-1}
  d\vect x_t = \Big(-\nabla V(\vect x_t) + E(t) D(\vect x_t)\Big)dt + \sqrt{2\beta^{-1}} d\vect w_t, 
\end{align}
where $x_t\in\Omega$ denotes the state of the molecular system at time $t$ in state space $\Omega$, $\vect w_t$ denotes standard $n$-dimensional Brownian motion, and $\beta$ the inverse temperature,
i.e.~$\beta = 1/(k_B\mathcal T)$.
\recheck{The Hamiltonain or Langevin dynamics can be treated in the same way as that will be presented, so for sake of simplicity we focus on the discussion of diffusive dynamics.}
The propagation of probability
densities $\rho=\rho(\vect x,t)$ based on this kind of dynamics in the sense
of $\rho(\vect x,t)dx=\prob[\vect x_t\in [\vect x,\vect x+d\vect x)]$ is governed by
Fokker-Planck equation:
\begin{align}
  \label{eq:disc-fp}
  \frac{\partial \rho}{\partial t} = \ml^\dagger(t) \rho,
\end{align}
where $\ml^\dagger(t)$ is the adjoint of the generator
\begin{align}
  \label{eq:disc-3}
  \ml(t)=\beta^{-1}\Delta_{\vect x}+\Big(-\nabla_x V(\vect x) + E(t)D(\vect x)\Big)\cdot\nabla_{\vect x},
\end{align}
where $\Delta_{\vect x}$ denotes the Laplacian operator and $\nabla_{\vect x}$
the nabla-operator wrt to $\vect x$. 
The periodicity of the external driving force induces the periodicity of the generator,
i.e.~$\ml(t) = \ml(t+T)$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial discretization: Master equation}
We will now introduce an appropriate spatial discretization of this kind of non-equilibrium MD -- \recheck{\bluec{(why saying this red sentence? we need the spatial discretization for calculating the core sets...)} this is done for reasons of simplicity only; we could completely avoid it for the price of more technical arguments.}
For achieving this discretization, we 
introduce a partition of state space $\Omega$ into a finite number of disjoint
sets $\{ \Omega_1, \cdots, \Omega_N\}$ satisfying $\Omega = \cup_i \Omega_i$,
$\Omega_j\cap \Omega_j = \emptyset,\ \forall i\neq j$.
Utilizing the procedure described in Ref.~\cite{latorre2011structure} the original Fokker-Planck equation~(\ref{eq:disc-fp})
is discretized, resulting in a time-inhomogeneous Markov jump process in state
space $S = \{1, \cdots, N\}$ with time-dependent rate
matrix $\vect L(t) \in \mathbb R^{N\times N}$ satisfying
\begin{align}\label{eq:disc-4}
\sum\limits_{j=1}^N L_{ij}(t) & =  0\\ \label{eq:disc-5}
L_{ij}(t) & \ge  0, \quad i\not= j\\
L_{ij}(t) & =  L_{ij}(t+T)
\end{align}
for all real time $t\geq 0$.
Moreover, the rate matrix $L$ has the form $\vect L(t)=\vect L_0+E(t)\vect L_1$
where $E(t)$ is periodic with period $T>0$.
In analogy to \eqref{eq:disc-fp}, the Markov jump process generated by
$\vect L(t)$ transports probability distributions according to the associated Master equation
\begin{align}
  \label{eq:disc-master}
  \frac{d\vect p(t)}{dt} = \vect L^{\top}(t)\cdot \vect p(t)
\end{align}
where $\vect L^{\top}(t)$ denotes the matrix transpose of $\vect L(t)$, $\vect p(t)$ is an $N$-vector denoting the probability distribution on $S$ at time $t$, $p(i,t)$, for example, the probability to be in state $i$ (which corresponds to set $\Omega_i$) at time $t$.
As usual the properties (\ref{eq:disc-4}) and (\ref{eq:disc-5}) of
$\vect L(t)$ guarantee that the total probability mass is conserved,
i.e., if $p(i,0)\ge 0$ componentwise, then $p(i,t)\ge 0$ and $\sum_i
p(i,t) = \sum_ip(i,0)$.
% The solution of the master equation need no be
% periodic.
The temporal evolution of the probability distribution $\vect p(t)$ can be formally written
\begin{align}  \label{eq:disc-8}
\vect p(t)=\myphi(t)\vect p(0)
\end{align}
by using the
associated propagator matrix $\myphi(t)\in\mathbb R^{N\times N}$ that solves
\begin{align}
  \label{eq:disc-master-phi}
  \frac{d}{dt}\myphi(t) = \vect L^{\top}(t)\myphi(t), \quad \myphi(0) = \id .
\end{align}
Since the last equation can be considered column-wise, the propagator matrix inherits column-wise conservation properties:
$\Phi_{ij}(t) \ge  0$
and $\sum\limits_{i=1}^N \Phi_{ij}(t)  =  1$,
that is, $\myphi^{\top}(t)$ is a
stochastic matrix satisfying $\myphi^{\top}(t)\vect e=\vect e$
with $\vect e=(1,\ldots,1)^{\top}\in \mathbb R^N$.
Regarding these considerations, we find
\begin{align}
\label{eq:disc-10}  
\myphi_{ij}(t)=\prob\left(\vect X_t=i\mid \vect X_0=j \right),
\end{align}
where $\vect X_t$ denotes the Markov process generated by $\vect L(t)$.

@@

Using the Floquet theorem, the time-inhomogeneous Markov jump process $X_t$
is therefore discretized into a \emph{time-homogeneous} (not necessarily
reversible) Markov jump process $\tilde X_{m} = X_{mT}, \ m\in\mathbb
N$, which is generated by transition matrix
\begin{equation}\label{P}
\vect P=\myphi^{\top}(T).
\end{equation}
We prefer to consider the discrete-time process $\tilde X_{m}$ instead of the time-continuous process $X_t$
because the powerful theories and
computational tools for time-homogeneous Markov processes can be directly applied. It worth noting that many of these tools require the transition matrix $P$ to satisfy the detailed balance condition. The computations in the Appendix~\ref{sec:app-revs} show that $P$ will in general \emph{not} satisfy this condition; in fact  the deviation {from} reversibility can be estimated from the work of the periodic driving does to the system.
There is no doubt that information within one period is lost by using this temporal
discretization, however, information regarding the long-term behavior of the system on timescales
much longer than the period will be perfectly described because of $\tilde X_{m} = X_{mT}\approx x_{mT}$ whenever our spatial discretization is fine enough.
At the same time,
the computational cost of generating $\tilde X_{m}$ is much less demanding
than the brute force simulations of NEMD, which implies lower
statistically uncertainty in calculating the observables of interest.

Since $P$ is a stochastic matrix, its eigenvalues are contained in the unit circle in
the complex plane, i.e., each eigenvalue $\lambda$ (potentially complex-valued)
satisfies $|\lambda|\le 1$. Furthermore $\lambda=1$ is an
eigenvalue with right eigenvector $\vect e=(1,\ldots,1)^{\top}$ and a left eigenvector $\mymu$
satisfying
$\mymu^{\top}P=\mymu^{\top}$.
From now on, we assume $P$ to be irreducible and aperiodic such that the Perron-Frobenius theorem \recheck{holds, so the eigenvector corresponding to} the eigenvalue $\lambda=1$ is  non-negative componentwise, and unique (up to normalization $\sum_i\mu(i)=1$). In this case $\mymu$ is the stationary measure in the sense that
$
 \mymu^{\top} P^m = \mymu^{\top},\ m\in\mathbb N
$,
% \begin{align}
% \label{eq:floq-14}  
% \end{align}
and (more precisely) the asymptotic evolution of an initial probability distribution $\vect p(t=0)$ by the process satisfies
$
\vect p^{\top}(0)\,P^m \to \mymu^{\top},\ m\to\infty,
$
% \begin{align}
% \label{eq:floq-15}  
% \end{align}
so that $\mymu$ can be seen as the quasi-stationary distribution of the non-stationary process.

% due to the discretization in both the spacial and temporal directions.
% In the numercal example in Sec.~\ref{sec:alanine}, we
% compute the

@@

% Given core sets $\{C_1, \cdots, C_k\}$, the procedure \eqref{eq:msm-ht}--\eqref{eq:msm-tmatrix-10}
% If assuming that we have the core sets $\{C_1, \cdots, C_k\}$, no matter how they are identified,

@@

We know that --in theory-- whenever the spatial discretization is fine enough, the Markov jump
process $X_t$ associated with the Master equation \eqref{eq:disc-master} is a good approximation to the original
MD process $x_t$ governed by \eqref{eq:disc-1}.  In practice, however, it is difficult to predict
how many discrete sets we need to be fine enough. Moreover, since the
total dimension of $x_t$ is $3N_{\textrm{atom}}$ ($N_{\textrm{atom}}$ being the number of atoms), it is prohibitive to do a really fine discretization
over all degrees of freedom for most systems of practical interest.

One possible way to define appropriate discretization sets is firstly to find a few collective
variables, and then to discretize these collective variables as finely as needed either by uniform or adaptive
discretization~\cite{chodera2007automatic, prinz2011markov}.
% This process also suggests that the choice of the lag-time should be 
% shorter than the dominant implied timescale so that they can be resolved,
% and be longer than the fast timescales to relax the unresolved dynamics
% within the discretized sets.
However, it is difficult to give a general answer in prior regarding
how to choose the collect variables and how fine their discretization should be.
For large or high dimensional systems, these questions usually become non-trivial.

@@

An important check is the consistency between the
stationary probability density of $\Phi(T)$ (i.e.~the leading eigenvector $\mymu$) and that 
estimated form the original NEMD simulation,
% along which only integral periods $mT$ are considered:
\begin{align}
  \label{eq:num-tmp1}
  \rho_{\textrm{st}}(\phi,\psi) = \lim_{m\rightarrow\infty} \rho (\phi,\psi,mT),
\end{align}
On each NEMD branching trajectory the initial 320~ps are discarded and
the rest of the trajectory in time interval $[320,4000]$~ps is \recheck{averaged} to estimate
the quasi-stationary probability distribution $\rho_{\textrm{st}}$. 
$P$ is computed as described above, and then $\mymu$ is computed as its leading eigenvector. In order to make it comparable to the free energy in the equilibrium case, we take
the logarithm of the distributions, i.e.~$F_{\textrm{st}}(\phi,\psi)=
-k_B\mathcal T\log \rho_{\textrm{st}}(\phi,\psi)$
for NEMD and $F_{\textrm{st}}(\phi,\psi)=
-k_B\mathcal T\log \mu(\phi,\psi)$ for $P$, where $k_B$ is the
Boltzmann constant and $\mathcal T$ is the temperature of the system.
% Under the aformentioned discretization, the stationary distribution is
% defined for the each bin of the dihedral angle space, i.e.~$F_{pq} =
% \int_{ph}^{(p+1)h}d\phi\int_{qh}^{(q+1)h}d\psi
% F_{\textrm{st}}(\phi,\psi), \ 0\leq p,q<N_\dih-1$, where $h = 360/N_\dih$ being
% the size of the bin, $p$ and $q$ here being the bin indexes.
The
results are  compared in Fig.~\ref{fig:num-1}. A good consistency between
the  NEMD simulation and 
$P$ is observed.
